---
title: "Impact of severe weather events from 1950 to 2011 in USA"
author: "Antonio Fernandez Vara"
date: "21 Mar 2015"
output: 
  html_document:
    keep_md: true
---

## Synopsis

Every year weather effects caused a significant amount of economic problems for communities and municipalities. The purpose of this anylysis was to calculate the damages for every type of weather event recorded for the years 1950 to 2011. All data for this analysis was obtained from the U.S. National Oceanic and Atmospheric Administration's storm database. This tracks events of mayour storms and weather events in the United States, with the time when they occur and an estimation of any fatalities, injuries and property damage.


## Data Processing

### Environment basic settings

```{r environment-basic-settings, echo=TRUE, cache=TRUE}
library(ggplot2)
library(xtable)
library(reshape2)
library(gridExtra)
```

### Obtaining data

This part of data processing focus on the download and extraction of the data to be able to replicate the analysis.

Data URL: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

The data is in CSV format compressed using bzip2.

```{r obtaining-data, echo=TRUE, cache=TRUE}
if (!file.exists('StormData.csv.bz2')) {
  url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
  filename <- 'StormData.csv.bz2'
  download.file(url, filename, method = 'curl')
}

if (!exists("storm.data")) {
  storm.data <- read.csv( bzfile('StormData.csv.bz2') )
}

if (!exists("raw.data")) {
  raw.data <- storm.data
}
```

Data must have 902297 observations (rows) of 37 variables (columns).

```{r dim-storm-data, echo=TRUE, cache=TRUE}
dim(storm.data)
```

Quick look of data

```{r head-storm-data, echo=TRUE, cache=TRUE}
head(storm.data)
```

### Cleaning data

Most important column to help categorize correctly the data is EVTYPE, but there are many different patterns for the same weather event. [Storm Database Documentation PDF](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) contains the documentation used to recategorizate to normalized events.

There are 985 different events recorded while the documentation of the database separate them into 48, so it needs to be cleaned and categorized using pattern matching. There are issues on how is labeled the data. 

NOTE: *Only significant events categories were cleaned, there are a more fine grane categorization to make still.*

```{r length-evtype, echo=TRUE, cache=TRUE}
length(unique(storm.data$EVTYPE))
```

To simplificate searchs, event type data is converted to capitals.

```{r toupper-evtype, echo=TRUE, cache=TRUE}
storm.data$EVTYPE <- toupper(as.character(storm.data$EVTYPE))
```

Normalizing data by pattern matching. Fixing some mispelled labels.

```{r normalizing-data, echo=TRUE, cache=TRUE}
storm.data[grep("ASTRONOMICAL", storm.data$EVTYPE),"EVTYPE"] <- "ASTRONOMICAL LOW TIDE"
storm.data[grep("AVALANCE|AVALANCHE", storm.data$EVTYPE),"EVTYPE"] <- "AVALANCHE"
storm.data[grep("BLIZZARD", storm.data$EVTYPE),"EVTYPE"] <- "BLIZZARD"
storm.data[grep("FLOOD|FLD|DROWNING|HIGH WATER|RISING WATER", storm.data$EVTYPE),"EVTYPE"] <- "COASTAL FLOOD"
storm.data[grep("COLD WIND CHILL|COLD/WIND CHILL|COLD/WINDS", storm.data$EVTYPE),"EVTYPE"] <- "COLD/WIND CHILL"
storm.data[grep("DEBRIS FLOW", storm.data$EVTYPE),"EVTYPE"] <- "DEBRIS FLOW"
storm.data[grep("FOG", storm.data$EVTYPE),"EVTYPE"] <- "DENSE FOG"
storm.data[grep("SMOKE", storm.data$EVTYPE),"EVTYPE"] <- "DENSE SMOKE"
storm.data[grep("DROUGHT|DRY", storm.data$EVTYPE),"EVTYPE"] <- "DROUGHT"
storm.data[grep("DUST DEVIL", storm.data$EVTYPE),"EVTYPE"] <- "DUST DEVIL"
storm.data[grep("DUST STORM", storm.data$EVTYPE),"EVTYPE"] <- "DUST STORM"
storm.data[grep("HEAT WAVE|EXCESSIVE HEAT|EXTREME HEAT|HEAT|WARM WEATHER", storm.data$EVTYPE),"EVTYPE"] <- "EXCESSIVE HEAT"
storm.data[grep("COLD|HYPOTHERMIA|LOW TEMPERATURE|EXPOSURE", storm.data$EVTYPE),"EVTYPE"] <- "EXTREME COLD/WIND CHILL"
storm.data[grep("FLASH FLOOD|FLOOD FLASH", storm.data$EVTYPE),"EVTYPE"] <- "FLASH FLOOD"
storm.data[grep("^FLOOD|FLD|DROWNING|HIGH WATER|RISING WATER", storm.data$EVTYPE),"EVTYPE"] <- "FLOOD"
storm.data[grep("^FROST|^FREEZE", storm.data$EVTYPE),"EVTYPE"] <- "FROST/FREEZE"
storm.data[grep("FUNNEL", storm.data$EVTYPE),"EVTYPE"] <- "FUNNEL CLOUD"
storm.data[grep("HAIL", storm.data$EVTYPE),"EVTYPE"] <- "HAIL"
storm.data[grep("HEAT|HIGH TEMPERATURE", storm.data$EVTYPE),"EVTYPE"] <- "HEAT"
storm.data[grep("RAIN|WET", storm.data$EVTYPE),"EVTYPE"] <- "HEAVY RAIN"
storm.data[grep("SNOW", storm.data$EVTYPE),"EVTYPE"] <- "HEAVY SNOW"
storm.data[grep("HIGH SURF", storm.data$EVTYPE),"EVTYPE"] <- "HIGH SURF"
storm.data[grep("HIGH WIND", storm.data$EVTYPE),"EVTYPE"] <- "HIGH WIND"
storm.data[grep("HURRICANE|TYPHOON", storm.data$EVTYPE),"EVTYPE"] <- "HURRICANE/TYPHOON"
storm.data[grep("ICE STORM|ICESTORM", storm.data$EVTYPE),"EVTYPE"] <- "ICE STORM"
storm.data[grep("LAKESHORE FLOOD|LAKE FLOOD", storm.data$EVTYPE),"EVTYPE"] <- "LAKESHORE FLOOD"
storm.data[grep("LAKE EFFECT SNOW|LAKE-EFFECT SNOW", storm.data$EVTYPE),"EVTYPE"] <- "LAKE-EFFECT SNOW"
storm.data[grep("LIGHTNING", storm.data$EVTYPE),"EVTYPE"] <- "LIGHTNING"
storm.data[grep("MARINE HAIL", storm.data$EVTYPE),"EVTYPE"] <- "MARINE HAIL"
storm.data[grep("MARINE HIGH WIND|MARINE STRONG WIND", storm.data$EVTYPE),"EVTYPE"] <- "MARINE HIGH WIND"
storm.data[grep("MARINE THUNDERSTORM WIND|MARINE TSTM WIND", storm.data$EVTYPE),"EVTYPE"] <- "MARINE THUNDERSTORM WIND"
storm.data[grep("RIP CURRENT", storm.data$EVTYPE),"EVTYPE"] <- "RIP CURRENT"
storm.data[grep("SEICHE", storm.data$EVTYPE),"EVTYPE"] <- "SEICHE"
storm.data[grep("SLEET", storm.data$EVTYPE),"EVTYPE"] <- "SLEET"
storm.data[grep("STORM SURGE", storm.data$EVTYPE),"EVTYPE"] <- "STORM SURGE/TIDE"
storm.data[grep("STRONG WIND|EXTREME WIND", storm.data$EVTYPE),"EVTYPE"] <- "STRONG WIND"
storm.data[grep("THUNDERSTORM|THUNDESTORM|THUNERSTORM|THUNDEERSTORM|THUNDEERSTORM|THUNDERTSORM|TSTM|TSTMW|WIND", storm.data$EVTYPE),"EVTYPE"] <- "THUNDERSTORM WIND"
storm.data[grep("TORNADO", storm.data$EVTYPE),"EVTYPE"] <- "TORNADO"
storm.data[grep("TROPICAL DEPRESSION", storm.data$EVTYPE),"EVTYPE"] <- "TROPICAL DEPRESSION"
storm.data[grep("TROPICAL STORM", storm.data$EVTYPE),"EVTYPE"] <- "TROPICAL STORM"
storm.data[grep("TSUNAMI", storm.data$EVTYPE),"EVTYPE"] <- "TSUNAMI"
storm.data[grep("VOLCANIC", storm.data$EVTYPE),"EVTYPE"] <- "VOLCANIC ASH"
storm.data[grep("WATERSPROUT|WATER SPROUT|WAYTERSPOUT", storm.data$EVTYPE),"EVTYPE"] <- "WATERSPROUT"
storm.data[grep("WILDFIRE|WILD FIRE", storm.data$EVTYPE),"EVTYPE"] <- "WILDFIRE"
storm.data[grep("WINTER STORM", storm.data$EVTYPE),"EVTYPE"] <- "WINTER STORM"
storm.data[grep("WINTER WEATHER|WINTRY|WINTERY", storm.data$EVTYPE),"EVTYPE"] <- "WINTER WEATHER"
```

### Filtered data

#### Creating a simpler dataset

For the calculations we need two subsets of the data we need to get our results. Also it calculates corrected economical damages using the exponential character as stated on the dataset documentation.

It takes three steps:

1. Calculate property damage corrected values using exponential column. It ignores all non documented exponents.


```{r simpler-dataset-1, echo=TRUE, cache=TRUE}
dataset <- storm.data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

for(i in 1:length(dataset$PROPDMG)){
    if(dataset$PROPDMGEXP[i] == "K" | dataset$PROPDMGEXP[i] == "k"){
        dataset$PROPDMG[i] <- dataset$PROPDMG[i] * 1000
    }
    if(dataset$PROPDMGEXP[i] == "M" | dataset$PROPDMGEXP[i] == "m"){
        dataset$PROPDMG[i] <- storm.data$PROPDMG[i] * 1000000
    }
    if(dataset$PROPDMGEXP[i] == "B" | dataset$PROPDMGEXP[i] == "b"){
        dataset$PROPDMG[i] <- dataset$PROPDMG[i] * 1000000000
    }
}
```

2. Calculate crop damage corrected values using exponential column. It ignores all non documented exponents.

```{r simpler-dataset-2, echo=TRUE, cache=TRUE}
for(i in 1:length(dataset$CROPDMG)){
    if(dataset$CROPDMGEXP[i] == "K" | dataset$CROPDMGEXP[i] == "k"){
        dataset$CROPDMG[i] <- dataset$CROPDMG[i] * 1000
    }
    if(dataset$CROPDMGEXP[i] == "M" | dataset$CROPDMGEXP[i] == "m"){
        dataset$CROPDMG[i] <- dataset$CROPDMG[i] * 1000000
    }
    if(dataset$CROPDMGEXP[i] == "B" | dataset$CROPDMGEXP[i] == "b"){
        dataset$CROPDMG[i] <- dataset$CROPDMG[i] * 1000000000
    }
}
```

3. Data for results with totals.

```{r simpler-dataset-3, echo=TRUE, cache=TRUE}
dataset.population <- dataset[, c("EVTYPE", "FATALITIES", "INJURIES")]
dataset.population <- melt(dataset.population, id = c("EVTYPE"))

dataset.economical <- dataset[, c("EVTYPE", "PROPDMG", "CROPDMG")]
dataset.economical <- melt(dataset.economical, id = c("EVTYPE"))
```


#### Most harmful with respect to population health.

Calculation of totals and sort by TOTAL column.

```{r totals-harmful-population, echo=TRUE, cache=TRUE}
dataset.population.damages <- dcast(dataset.population, EVTYPE ~ variable, sum)
dataset.population.damages$TOTAL <- rowSums(dataset.population.damages[, c("FATALITIES", "INJURIES")])
dataset.population.damages <- dataset.population.damages[order( -dataset.population.damages[,4] ), ]
``` 


#### Most economical damage totals.

Calculation of totals and sort by TOTAL column.

```{r totals-economical-damages, echo=TRUE, cache=TRUE}
dataset.economical.damages <- dcast(dataset.economical, EVTYPE ~ variable, sum)
dataset.economical.damages$TOTAL <- rowSums(dataset.economical.damages[, c("PROPDMG", "CROPDMG")])
dataset.economical.damages <- dataset.economical.damages[order( -dataset.economical.damages[,4] ), ]
```


## Results

### Most Harmful types of events with respect to population health.

The most harmful events are in order tornados, thunderstorm winds and heat. Tornados the most harmful event for both fatalities and injuries.

The table below shows the top 10 events that have the greatest impact on population health.

```{r table-harmful-population, echo=TRUE, cache=TRUE}
  head(dataset.population.damages, n=10)
```

Plot with those values.

```{r plot-harmful-population, echo=TRUE, cache=TRUE}
  ggplot(melt(head(dataset.population.damages[,c(1:3)], n=10)), 
          aes(x = EVTYPE, y = value, fill = variable)) +
          geom_bar(stat="identity", width = 0.7) +
          xlab("Recorded Events") + 
          ylab("Total Fatalities and Injuries") +
          ggtitle("Population damaged by weather events") +
          scale_fill_hue(name = "Damage Type") + 
          theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Most economical impact on property and crop damage.

The most impacting events globally are coastal floods, hurricanes and tornados. Properties are impacted by the action of coastal flood, while crops get affected by drought mainly.

The table below shows the top 10 events that have the greatest impact on property, crop and global damage.

```{r table-economical-damage, echo=TRUE, cache=TRUE}
  head(dataset.economical.damages, n=10)
```

Plot showing the values.

```{r plot-economical-damage, echo=TRUE, cache=TRUE}
  ggplot(melt(head(dataset.economical.damages[,c(1:3)], n=10)), 
          aes(x = EVTYPE, y = value, fill = variable)) +
          geom_bar(stat="identity", width = 0.7) +
          xlab("Recorded Events") + 
          ylab("Total Properties and Crop damages") +
          ggtitle("Economical impact of severe weather events") +
          scale_fill_hue(name = "Damage Type") + 
          theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
